To be FAIR: Theory Specification Needs an Update

### 2025-03-05 manuscript.Rmd
Final tweaks

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, the Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, Germany"
  - id            : "7"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`. See the CRediT authorship contributions and GitHub repository for specific author contributions.

@@

The so-called "replication crisis" has prompted extensive reforms in psychology and other scientific fields [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The profound difference between these definitions becomes clear when considering the act of changing a child’s dirty diaper.
Under the original definition, few caregivers would be considered intrinsically motivated, after all, it’s not exactly a joy-filled experience.
Yet, under the expanded definition, many would be, as the act may fulfill deeper needs, such as the desire to nurture and care for one’s child.

<!-- CJ: Maybe we should give an example here. Does anyone know one? -->
<!-- NvD: Harris 1976 has a nice example about how cognitive dissonance theory can be formalized in multiple ways, which offer hypotheses that contradict each other -->

Scholars have been raising concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories are difficult to understand on their own,
without either substantial interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory. 
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and improvement have been largely overlooked^[Here, we use "induction" to refer to inferences from specific observations to general theories. Others have used the term "abduction", coined by Peirce to describe "inference to the best explanation". For present purposes, however, the terms are interchangeable.].
<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences. Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].  NvD dissents: -->
<!-- NvD: I cannot accept this footnote as an adequate characterization. Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more). See for instance Brian Haig's 2005 paper. Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions". At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree. I am fine with the last sentence though ;-) My suggestion is to write the following: -->
<!-- NvD: We use induction to describe inferences from specific observations to general theories. In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations. We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable. -->
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are thoughtlessly reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978].

@@

The present paper introduces open science methods for theory specification and archival.
Merely publishing theory in a research article does not make it open;
to be open, theory should adhere to established open science standards.
We do so by implementing theories as information artifacts,
archived with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data", they provide information about the nature and content of an information artifact.
Metadata are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal.
At the same time, formal theories are not automatically FAIR.
The FAIR principles pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility.
They can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).

@@

We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable *information artifacts* in their own right,
with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
and cross-reference the published paper with `relationType: IsSupplementTo`.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`. Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<!-- FS: I still think we should have a dedicated term for a digital FAIR theory object -->

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013].
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
Note, however, that the function of Git(Hub) is *not* to archive a FAIR theory.
FAIR-compliant repositories like Zenodo serve the specific purpose of archival.
Thus, while theory development may take place on GitHub, the version of record should be archived in Zenodo, with appropriate metadata.

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and post many examples on their website [@grayHowMapTheory2017].
Similarly, Van Dongen and colleagues are working on a database of models and formalized theories.
<!-- Is there something to reference? Website/preprint? Otherwise is seems a bit weird to just mention this without any further information. -->
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016a].
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage2024]
for representing graphs does meet this ideal (an example of such a DOT representation is given below).

@@

By default, Zenodo assumes that GitHub repositories contain software and documents them as such.
To document our archive as a FAIR theory requires adding some extra information on Zenodo.
Supplying the information below helps improve the Findability of a theory.
See here for [an example of the resulting FAIR metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) archived in Zenodo.

>>>>

author:
  - name: "Caspar J. Van Lissa"
    affiliation: "1"
    corresponding: yes
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Formal Analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name: "Aaron Peikert"
    affiliation: "2,3"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
  - name: "Maximilian S. Ernst"
    affiliation: "2,4"
    role:
      - Writing – review & editing
      - Investigation
  - name: "Noah N.N. van Dongen"
    affiliation: "5"
    role:
      - Writing – review & editing
  - name: "Felix D. Schönbrodt"
    affiliation: "6"
    role:
      - Conceptualization
      - Writing – review & editing
  - name: "Andreas M. Brandmaier"
    affiliation: "2,3,7"
    role:
      - Formal Analysis
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, dept. Methodology & Statistics"
  - id            : "2"
    institution   : "Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany"
  - id            : "3"
    institution   : "Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany"
  - id            : "4"
    institution   : "Max Planck School of Cognition, Leipzig, Germany"
  - id            : "5"
    institution   : "University of Amsterdam, the Netherlands"
  - id            : "6"
    institution   : "Ludwig-Maximilians-Universität München, Germany"
  - id            : "7"
    institution   : "Department of Psychology, MSB Medical School Berlin, Berlin, Germany"
authornote: |
  This is a preprint paper, generated from Git Commit # `r substr(gert::git_commit_id(),1,8)`. The first and senior author positions were fixed, and remaining author positions randomized using `set.seed(6999982); sample(c("Aaron", "Felix", "Noah", "Maximilian"))`.

@@

The so-called "replication crisis" has prompted extensive scientific reforms [@lavelleWhenCrisisBecomes2021; @scheelWhyMostPsychological2022].
Concern that the abundance of non-replicable findings 
was caused by undisclosed flexibility in analyses led to widespread adoption of open science practices like preregistration and replication [@nosekPromotingOpenResearch2015a].
These various practices ensure transparent and repeated testing of hypotheses by commiting to an analysis plan in advance.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence [@scheelExcessPositiveResults2021].
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.
A good example of this is found in “self-determination theory” (SDT), which emphasizes the role of intrinsic and extrinsic motivation in human behavior.
Initially, intrinsic motivation was understood as engaging in an activity purely for the inherent satisfaction it provides, free from any external rewards or pressures [@deciEffectsExternallyMediated1971].
Over time, however, SDT expanded its definition to include motivations driven by the fulfillment of basic psychological needs—autonomy, competence, and relatedness—all still categorized as "intrinsic" [@ryanSelfdeterminationTheoryFacilitation2000].
The difference between these definitions becomes clear when considering the act of changing an infant's dirty diaper.
Under the original definition, one might hypothesize that caregivers are not intrinsically motivated to change diapers, as this is hardly a joyous experience.
Under the expanded definition, one might hypothesize the opposite, as the act may fulfill the need for relatedness.

Scholars have raised concerns about the state of theory in psychology for nearly 50 years [@robinaughInvisibleHandsFine2021; @meehlTheoreticalRisksTabular1978].
One main concern is that theories lack *formalization* [@szollosiArrestedTheoryDevelopment2021].
<!-- I still believe formalization != precision. Theories can be 100% formal and still 0% precise. -->
<!-- NvD: We should add somewhere what we mean by "formalization." In philosophy of science, formal logic about syntax (the propositions and predicates can have any meaning). However, what we mean, I think, is the explication of a narrative in (something so detailed and explicit that it can be captured with little error in) analytical or computational form (but the meaning remains. E, M, and C stand form something is E = MC<sup>2</sup>).   -->
When theories are difficult to understand without either subjective interpretation or additional background knowledge,
it becomes difficult do derive precise predictions,
and therefore hard to falsify the theory. 
A second concern is the lack of transparent and participative scholarly communication about psychological theory, which limits its progression and development.
Despite these concerns, scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and improvement have been largely overlooked^[Here, we use "induction" to refer to inferences from specific observations to general theories. Others have used the term "abduction" to describe "inference to the best explanation" [@peirceCollectedPapersCharles1960]. For present purposes, the terms are interchangeable.].
<!-- CvL considers abduction to be a special case of induction, because defining what makes a theory "best" or "plausible" requires making auxiliary assumptions, which restrict the space of potential inductive inferences. Peirce also stated that abduction "is not essentially different from induction" [@peirceCollectedPapersCharles1960, p.4881].  NvD dissents: -->
<!-- NvD: I cannot accept this footnote as an adequate characterization. Abduction is not just "inference to the best explanation." It is typically used as inference to a plausible explanation (but there could be more). See for instance Brian Haig's 2005 paper. Also, I don't see why it matters either way "why defining what makes a theory "best" requires auxiliary assumptions". At the very least I would like to change "we consider abduction to be a special case of induction", because I disagree. I am fine with the last sentence though ;-) My suggestion is to write the following: -->
<!-- NvD: We use induction to describe inferences from specific observations to general theories. In philosophy of science the term "abduction" is often used for the inference from observations to plausible explanations. We note that induction and abduction come in several varieties in philosophy science, which overlap in several respects For present purposes, however, the terms are interchangeable. -->
<!-- Peirce: CP 8.233 Cross-Ref:†† -->
<!-- 233. This is not essentially different from induction. Only it is most -->
<!-- usually an induction from instances which are not discrete and numerable. I now -->
<!-- call it Qualitative Induction. It is this which I used to confound with the second -->
<!-- line of procedure, or at least not to distinguish it sharply. -->
The present paper addresses this knowledge gap by applying, for the first time,
open science principles to psychological theory.
We apply introduce *FAIR theory* to 
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

In a progressive research program [@lakatosHistoryScienceIts1971],
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena.
<!-- NvD: This is not exactly what Lakatos said. Also, I don't think many psychologists are familiar with his work. We do not need to reference his perspective. -->
<!-- by successively increasing the empirical content of a theory. -->
<!-- In psychology, this is typically achieved by either generalizing to a larger set of people, contexts, situations, or environments. <!--AB would you agree?-->
<!-- CJ: I am not sure if I agree; I think increasing the empirical content rather implies more and more precision, which loses generality but gains specificity. Not sure how to resolve.
This needs more work! -->
<!-- NvD: SOmething like? "In the envisioned ideal, the empirical cycle is a virtuous spiral where theories become increasingly more precise and/or universal (i.e., Popper's empirical content), offering hypotheses that can be more easily falsified / severely tested, data become more informative / less ambiguous about the patterns in nature, which in turn inform even more precise and/or universal theories. -->
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature: 
According to @kuhbergerPublicationBiasPsychology2014, 89.6% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of deductive research reveals, however, that the link between theory and hypothesis is often tenuous or absent [@scheelWhyHypothesisTesters2021; @oberauerAddressingTheoryCrisis2019].
Only 15% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis [@mcphetresDecadeTheoryReflected2021].
The remaining 85% of deductive studies lacked an explicit connection between theory and hypothesis.
Perhaps some of these ungrounded hypotheses are rooted in implicit theories privy only to the author,
in which case it would be straightforward and important to make these explicit [@friedTheoriesModelsWhat2020; @norouziCapturingCausalClaims2024].
Or, perhaps some hypotheses are reported as part of entrenched research practices [@gigerenzerNullRitualWhat2004],
but are not of substantive interest, such as null hypotheses that exist solely for the purpose of being rejected [@vanlissaTeacherCornerEvaluating2020].
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
<!-- NvD: I don't think that statement completely covers it. How about: "Testing ad-hoc hypotheses not grounded in theory might inform us of the psychological phenomena that we can consistently observe (e.g., cognitive biases, attentional blindess, comorbidity of depression and anxiety disorder, etc.). However, it does not advance our understanding of the processes and mechanisms that bring about these phenomena. -->
<!-- CJ: I agree, but the point about phenomena detection is not directly relevant to this paper, so I prefer not to get into it here. -->
Collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings [@vanrooijTheoryTestHow2021].

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected to, let alone derived from, theory.
Moreover, test results do not routinely contribute to the improvement or rejection of theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology "lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest" [@meehlTheoreticalRisksTabular1978].
The present paper seeks to make theory more tangible and practical by instantiating it as a digital "object" that scholars can access, reuse, and update in their daily workflows.

@@

Merely publishing theory in a journal article does not make it open;
to be open, theory should adhere to established open science standards for specification and archival.
We propose to implement theories as information artifacts,
and archive these with appropriate metadata in a FAIR-compliant repository (e.g., Zenodo).
Metadata are "data about the data". 
They provide information about the nature and content of an information artifact and are stored in the repository where the version of record of the FAIR theory is deposited.
FAIR theories are *Findable* via a DOI or by searching the repository they are archived in;
*Accessible* in a machine- and human-readable filetype;
*Interoperable* for specific purposes, for example, within the data analysis environment;
and *Reusable* in the practical and legal sense, so that they may be iteratively improved by the author or by others.
<!-- AB I think some initial word on what is covered is needed here; otherwise readers may wonder what we mean to cover--> 
<!-- CJ: Since this is in the next paragraph, I think it's redundant. Will merge there -->
Following the original proposal of Lamprecht and colleagues,
we adapt the FAIR principles for theory, see [Supplemental Table S1](https://github.com/cjvanlissa/fair_theory/blob/main/fair_principles.csv).
We reflect on the necessary changes (which are minor),
as well as on the current state and future of FAIR theory in psychology.
The resulting principles provide guidance for instantiating theory as a FAIR information artifact,
and we provide worked examples to encourage their adoption.

@@

FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
Thus, FAIR theory does not require theories to be formal.
At the same time, formal theories are not automatically FAIR.
The FAIR principles
<!-- pertain to theories' metadata documentation and sharing in digital archives, with the aim of enhancing their reusability and extensibility. -->
<!-- They  -->
can be applied to theories representated in natural language,
as well as formal theories represented using mathematical notation, algorithmic pseudo code, or a set of logical clauses.
Thus, for example, "grounded theory", derived from qualitative research,
can be represented as a FAIR theory if it is represented as plain-text propositions and archived in a FAIR repository with appropriate metadata.
Conversely, a formal theory is not FAIR if it is confined to a journal article without any key words to identify it as a theory paper (lacking Findability), represented merely as a bitmap image (limiting Accessibility and Interoperability), or behind a paywall (limiting Reusability).
FAIR theory is thus consistent with, but does not require, formalization (also see the section on *Accessibility* below).

@@

We propose FAIR theory as an instantiation of modular publishing [@kirczModularityNextForm1998].
In most fields, the primary unit of scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable *information artifacts*,
with adequate metadata that is indexed in standardized repositories [@vandesompelRethinkingScholarlyCommunication2004].
Data sharing is an example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like [Zenodo](https://zenodo.org/),
which was developed by [CERN](https://home.cern/) under the European Union's [OpenAIRE](https://www.openaire.eu/) program [@zenodo].
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the [Crossref](https://www.crossref.org/) association,
which is used by many academic publishers.
Finally, the DataCite Metadata Schema offers a standard way to document the nature of relationships between information artifacts [@datacitemetadataworkinggroupDataCiteMetadataSchema2024].
<!-- Platforms like the [Open Science Framework](https://osf.io/) connect these various research infrastructures to offer a more user-friendly front end for open science-related workflows. -->
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property `resourceType: dataset`,
and cross-reference the published paper with `relationType: IsSupplementTo`.
Similarly, a FAIR theory object can be connected to a specific paper which might serve as the theory's documentation and canonical reference by using `relationType: IsDescribedBy`, while the reverse relationship, documented in the canonical reference paper, is `relationType: Describes`. Other types are useful for making relationships between multiple theory objects explicit: If an existing theory is made FAIR without substantial alterations,
the resulting FAIR theory metadata would cross-reference the existing theory as `relationType: IsDerivedFrom`.
If an existing theory is updated, `relationType:IsNewVersionOf` could be used to reference previous versions.
If a variation of an existing FAIR theory is created, cross-reference it with `relationType: IsVariantFormOf`.
<!-- FS: I still think we should have a dedicated term for a digital FAIR theory object -->

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions from collaborators and allowing for experimentation and diverging development without losing information.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work [@ramGitCanFacilitate2013; @vanlissaWORCSWorkflowOpen2021].
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration [@vanlissaWORCSWorkflowOpen2021; @peikertReproducibleResearchTutorial2021].
Git provides a useful framework for developing FAIR theory,
because it enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].
Note that archival of the version of record is *not* a function of Git(Hub).
While theory development may take place on GitHub, the version of record should be archived in a FAIR-compliant archive like Zenodo, with appropriate metadata.

@@

There have been notable efforts to improve theories' Findability through post-hoc curation.
For example, Gray and colleagues introduced a format for representing theories,
and posted many examples on their website <www.theorymaps.org> [@grayHowMapTheory2017].
Similarly, [PsychoModels](http://psychomodels.org/) seeks to inventorize theories and models in psychology [@vandongenPsychoModelsMathematicalComputational2025].
Post-hoc curation is a notable effort but does not address the root cause of the lack of Findability.
Ideally, Findability would be addressed ante-hoc, through documentation with rich metadata and modular publishing.
Both approaches can be complementary, however.
For example, post-hoc curation could make use of existing FAIR-compliant archival infrastructure like Zenodo.
Conversely, the database engineering adage "Lots of Copies Keeps Stuff Safe" (LOCKSS) implies that it is fine to archive theories in multiple places,
although it is advisable to make use of automatic integration (as exists between GitHub, Zenodo, and OSF) to avoid the need to maintain information in multiple places, which increases the risk of inconsistencies arising.

@@

Interoperability pertains to the property of information artifacts to "integrate or work together [...] with minimal effort" [@wilkinsonFAIRGuidingPrinciples2016].
Firstly, theory and its associated metadata should use a formal, accessible, shared and broadly applicable language to facilitate (human- and) machine readability and reuse (I1).
The common practice of instantiating theory as lengthy prose or schematic drawing falls short of this ideal.
Instead, FAIR theory should, ad minimum,
be instantiated in a human- and machine-readable datatype,
as should all information artifacts created while performing scholarly work [@vanlissaWORCSWorkflowOpen2021].
Depending on the level of formalization of the theory,
different formats may be appropriate,
such as verbal statements in plain text,
mathematical formulae,
and statements expressed in some formal language.
Examples of the latter include pseudo-code,
interpretable computer code,
and Gray's theory maps [@grayHowMapTheory2017].
While a theory represented as a bitmap image is not very Interoperable,
the same image represented in the DOT language [@DOTLanguage2024]
for representing graphs does meet this ideal (an example of such a DOT representation is given below).

@@

To document our archive as a FAIR theory and improve its Findability requires adding the relevant metadata on Zenodo.
See here for [an example of the metadata](https://raw.githubusercontent.com/cjvanlissa/fair_theory/refs/heads/main/example_metadata.json) associated with our FAIR empirical cycle, as archived in Zenodo.
